Approximately Exact Calculations for Linear Mixed Models

### 2015-06-02 ms.tex
mostly Andrew's edits

We work with one term in (\ref{eq:partials})'s summations at a time; that is, one $j$ at a time.  For $j=1, \dots, s_z$, the $j$'th terms in (\ref{eq:partials}) differ  by a multiplicative constant $a_j/b_j = a_j$; they have the same sign as each other and the same sign as $(a_jc_j\sigssq + b_jc_j\sigesq  - d_j) = (a_j\sigssq + \sigesq  - d_j)$, which determines a line $\sigssq = d_j/a_j - \sigesq/a_j$ --- call it the $j$'th line --- in the first quadrant of the $(\sigssq,\sigesq)$ plane.  The $j$'th line has positive intercept $ d_j/a_j$ and negative slope $-1/a_j$.  For $j=s_z+1$, $a_j=0$, so $\eqref{eq:drlds} = 0$ and \eqref{eq:partials} determines a vertical line at $\sigesq = \sigma_e^{2*} \equiv d_{s_z+1} / c_{s_z+1}$.  For $j=s_z+2$,  $b_j=0$, so $\eqref{eq:drlde} = 0$ and, if the target function is \eqref{eq:logpost}, \eqref{eq:partials} determines a horizontal line at $\sigssq = \sigma_s^{2*} \equiv d_{s_z+2} / c_{s_z+2}$, while if the target function is \eqref{eq:reexpress}, $c_{s_z+2} = d_{s_z+2} = 0$; the term for $j=s_z+2$ effectively vanishes, and there is no horizontal line.

@@

Let $\sigesq{}^M$ and $\sigssq{}^M$ be the largest intercepts of the $s_z+2$ lines on the $\sigesq$ and $\sigssq$ axes, respectively.  I.e.,
\begin{align*}
  \sigesq{}^M &= \max \left\{ \frac{d_1}{c_1}, \cdots, \frac{d_{s_z+1}}{c_{s_z+1}} \right\}\\
  \sigssq{}^M &= \begin{cases}
                             \max \left\{ \frac{d_1}{a_1}, \cdots, \frac{d_{s_z}}{a_{s_z}} \right\} &
                                  \text{if $\logRL$ is \eqref{eq:reexpress}}\\
                             \max \left\{ \frac{d_1}{a_1}, \cdots, \frac{d_{s_z}}{a_{s_z}}, \frac{d_{s_z+2}}{c_{s_z+2}} \right\} &
                                  \text{if $\logRL$ is \eqref{eq:logpost}}.
                           \end{cases}
\end{align*}
Let $B_1$ be the box whose lower-left and upper-right corners are $(0,0)$ and $(\sigesq{}^M,\sigssq{}^M)$, respectively.  Figure~(\ref{fig:bigboundingbox}) illustrates with five lines --- realistic data sets may have more --- labelled $j=1, 2, 3, s_z+1, s_z+2$.\\
\textbf{Claim} \emph{All local maxima of $\logRL$ lie within $B_1$.}\\
\textbf{Proof} Let $p_1 = (\sigesq{}_1, \sigssq{}_1)$ be a point such that $\sigssq{}_1 > \sigssq{}^M$ and let $\widetilde{p_1} = (\sigesq{}_1, \sigssq{}^M)$, as illustrated in Figure~(\ref{fig:bigboundingbox}).  For $j=1, \dots, s_z, s_z+2$, the partial derivatives \eqref{eq:drlde} are negative everywhere between $p_1$ and $\widetilde{p_1}$; for $j=s_z+1$, the partial derivative is zero.  Therefore, $\logRL(\widetilde{p_1}) \ge \logRL(p_1)$ and there can be no local maxima of $\logRL$ above the line $\sigssq = \sigssq{}^M$.  Let $p_2 = (\sigesq{}_2, \sigssq{}_2)$ be a point such that $\sigesq{}_2 > \sigesq{}^M$ and let $\widetilde{p_2} = (\sigesq{}^M, \sigssq{}_2)$.  For $j=1, \dots, s_z, s_z+1$, the partial derivatives \eqref{eq:drlds} are negative everywhere between $p_2$ and $\widetilde{p_2}$; for $j=s_z+2$, the partial derivative is zero.   Therefore, $\logRL(\widetilde{p_2}) \ge \logRL(p_2)$ and there can be no local maxima of $\logRL$ to the right of the line $\sigesq = \sigesq{}^M$.  \textbf{QED}\\

By the claim, \mrle\ must lie on or inside $B_1$, so $B_1$ could be passed to an optimizer such as \R's \textsf{optim} or \textsf{nlminb} with potentially better results than using those functions without bounds.  However, even with known bounds, general purpose optimizers may still miss \mrle\ and regions of high $\logRL$ and they emphasize single points of highest local $\logRL$ while possibly ignoring large adjacent regions where $\logRL$ is nearly as large \citep{hill:1965, hodges:2013}.  We will next find a box $B \supset B_1$ satisfying desideratum \textbf{D1}.  First, though, we pause to note that the region outlined in bold in Figure~(\ref{fig:smallboundingbox}) is a subset of $B_1$ that must also contain \mrle\ by the same reasoning used to prove the previous claim.  But it is not rectangular, hence less convenient than $B_1$, so we don't pursue it further.

@@

@@

Let $q_2 = (\sigesq{}_2, \sigssq{}_2)$ satisfy $\sigesq{}_2 \ge \sigma_e^{2*}$ and $\sigssq{}_2 \ge \widetilde{\sigma_s^2}$ --- above $B$ and to the right of the $s_z+1$ line.  (Points $q_2$ and $q_3$ in Figure~(\ref{fig:boundingbox2}) both satisfy this criterion.)  The partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) evaluated at $q_2$ are nonpositive for $j \in \{1, \dots, s_z+2\}$; therefore, $\logRL(\sshat) - \logRL(q_2) \ge L - \logRL(\widetilde{q_2})$.  The latter expression is greater than $M$ iff $\logRL(\widetilde{q_2}) < L-M$.  But examination of (\ref{eq:sum}) shows that for any fixed $\sigesq$, $\lim_{\sigssq \rightarrow \infty} \logRLss = -\infty$ and therefore $\widetilde{\sigma_s^2}$ can be chosen large enough so that \textbf{D1} is satisfied.

@@

\subsection{Lines Determine Bounds within Boxes}
\label{sec:boundsinboxes}
\begin{figure}
	\centering
	\includegraphics[width=.5\linewidth]{figs/boxb.pdf}
	\caption{For a box $b$,  the minimum and
	maximum within $b$ of $\logRL_j$
	occur at either the lower left corner, the upper right  corner, or
	on the line.
}	\label{fig:boxb}
\end{figure}
Next we consider the relationship between the $\logRL_j$'s and boxes as depicted in Figure~\ref{fig:boxb}, which shows a box $b$ and three lines.  For lines such as $j=2$ that lie below or to the left of $b$, the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are nonpositive, so the maximum and minimum of $\logRL_j$ within $b$ are attained at the lower left and upper right corners respectively.  The situation is reversed for lines like $j=3$ that lie above or to the right of $b$: the partial derivatives are nonnegative so the maximum and minimum are attained at the upper right and lower left corners respectively.  For lines like $j=1$ that pass through $b$, the minimum is attained at either the upper right or lower left corner while the maximum is attained on the line.

@@

\section{Examples}

\subsection{HMO premiums}
\subsubsection{Introduction to the Data}
Our first example is a traditional linear mixed model previously analyzed in \cite{hodges:98}, \cite{hodges:2013}, and \cite{henn&hodges:2014} who reported a bimodal log posterior density for $(\sigesq,\sigssq)$.  Quoting from \cite{henn&hodges:2014},
\begin{quote}
\itshape
\dots the HMO data set describes 341 HMOs [Health Maintenance Organizations] located in 45 states or similar political jurisdictions.  Each jurisdiction had between 1 and 31 plans with a median of 5 plans.  The data set originally was analysed to assess the cost of moving military retirees and dependents from a Department of Defense health plan to plans serving the US civil service.  \dots.  Specifically, the model is
\begin{equation*}
\begin{split}
	y_{ij} &= \alpha_i + \epsilon_{ij}\\
	\alpha_i &= \varrho_0 + \varrho_1x_{1i} + \varrho_2x_{2i} + \zeta_i,
\end{split}
\end{equation*}
where the fixed effects in $\alpha_i$ include an intercept, jurisdiction-average hospital expenses per admission ($x_{1i}$) and an indicator for plans in New England states ($x_{2i}$).

@@

Figure~\ref{fig:hmoboxes} shows the outlines of the 9490 boxes.  The algorithm did not need to divide the boxes with large $\sigesq$ or $\sigssq$ as finely as those with small $\sigesq$ and $\sigssq$ because they more readily satisfy either $U^b < L-M$ or $U^b - L^b < \epsilon$, so become inactive.  Figure~\ref{fig:hmorll} shows the same boxes, shaded by $L^b$ in each box.  The red dot is the lower left corner of the box that maximizes $L^b$.  Boxes with $U^b \ge L \equiv \max L^b$ are outlined in blue; $(\sshat)$ must lie within the blue region.  
%Figure~\ref{fig:hmorll}  suggests that we can restrict attention to a box determined by $\sigesq\in(300,1000)$ and $\sigssq\in(0,1000)$.  So we rerun the algorithm on that box and with more stringent control parameters:
%\begin{equation*}
%	\maxit=15; \hspace{1cm} \epsilon=1; \hspace{1cm}
%	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=10
%\end{equation*}
%Table~\ref{table:hmo_HH11_run2} shows the output.  The algorithm needed 12 iterations to move all boxes to the inactive list.  The resulting 37,516 boxes are shown in Figure~\ref{fig:hmo2}.
For comparison, the standard REML analysis using \R's \textcompute{lme} function yields $\sshat \approx (495, 99)$ with 95\% confidence intervals of $(421, 582)$ and $(39, 248)$.

>>>>

We work with one term in (\ref{eq:partials})'s summations at a time; that is, one $j$ at a time.  For $j=1, \dots, s_z$, the $j$'th terms in (\ref{eq:partials}) differ  by a multiplicative constant $a_j/b_j = a_j$; they have the same sign as each other and the same sign as $(a_jc_j\sigssq + b_jc_j\sigesq  - d_j) = (a_j\sigssq + \sigesq  - d_j)$, which determines a line $\sigssq = d_j/a_j - \sigesq/a_j$ --- call it the $j$'th line --- in the first quadrant of the $(\sigssq,\sigesq)$ plane.  The $j$'th line has positive intercept $ d_j/a_j$ and negative slope $-1/a_j$.  For $j=s_z+1$, $a_j=0$, so $\eqref{eq:drlds} = 0$ and \eqref{eq:partials} determines a vertical line at $\sigesq = \sigma_e^{2*} \equiv d_{s_z+1} / c_{s_z+1}$.  For $j=s_z+2$,  $b_j=0$, so $\eqref{eq:drlde} = 0$ and, if the target function is \eqref{eq:logpost}, \eqref{eq:partials} determines a horizontal line at $\sigssq = \sigma_s^{2*} \equiv d_{s_z+2} / c_{s_z+2}$, while if the target function is \eqref{eq:reexpress}, $c_{s_z+2} = d_{s_z+2} = 0$; the term for $j=s_z+2$ effectively vanishes and there is no horizontal line.

@@

Let $\sigesq{}^M$ and $\sigssq{}^M$ be the largest intercepts of the $s_z+2$ lines on the $\sigesq$ and $\sigssq$ axes, respectively.  I.e.,
\begin{align*}
  \sigesq{}^M &= \max \left\{ \frac{d_1}{c_1}, \cdots, \frac{d_{s_z+1}}{c_{s_z+1}} \right\}\\
  \sigssq{}^M &= \begin{cases}
                             \max \left\{ \frac{d_1}{a_1}, \cdots, \frac{d_{s_z}}{a_{s_z}} \right\} &
                                  \text{if $\logRL$ is \eqref{eq:reexpress}}\\
                             \max \left\{ \frac{d_1}{a_1}, \cdots, \frac{d_{s_z}}{a_{s_z}}, \frac{d_{s_z+2}}{c_{s_z+2}} \right\} &
                                  \text{if $\logRL$ is \eqref{eq:logpost}}.
                           \end{cases}
\end{align*}
Let $B_1$ be the box whose lower-left and upper-right corners are $(0,0)$ and $(\sigesq{}^M,\sigssq{}^M)$, respectively.  Figure~(\ref{fig:bigboundingbox}) shows $B_1$ and five lines --- realistic data sets may have more --- labelled $j=1, 2, 3, s_z+1, s_z+2$.\\[5pt]
\textbf{Claim} \emph{All local maxima of $\logRL$ lie within $B_1$.}\\
\textbf{Proof} Let $p_1 = (\sigesq{}_1, \sigssq{}_1)$ be a point such that $\sigssq{}_1 > \sigssq{}^M$ and let $\widetilde{p_1} = (\sigesq{}_1, \sigssq{}^M)$, as illustrated in Figure~(\ref{fig:bigboundingbox}).  For $j=1, \dots, s_z, s_z+2$, the partial derivatives \eqref{eq:drlde} are negative everywhere between $p_1$ and $\widetilde{p_1}$; for $j=s_z+1$, the partial derivative is zero.  Therefore, $\logRL(\widetilde{p_1}) \ge \logRL(p_1)$ and there can be no local maxima of $\logRL$ above the line $\sigssq = \sigssq{}^M$.  Let $p_2 = (\sigesq{}_2, \sigssq{}_2)$ be a point such that $\sigesq{}_2 > \sigesq{}^M$ and let $\widetilde{p_2} = (\sigesq{}^M, \sigssq{}_2)$.  For $j=1, \dots, s_z, s_z+1$, the partial derivatives \eqref{eq:drlds} are negative everywhere between $p_2$ and $\widetilde{p_2}$; for $j=s_z+2$, the partial derivative is zero.   Therefore, $\logRL(\widetilde{p_2}) \ge \logRL(p_2)$ and there can be no local maxima of $\logRL$ to the right of the line $\sigesq = \sigesq{}^M$.  \textbf{QED}\\

@@

By the claim, \mrle\ must lie on or inside $B_1$, so $B_1$ could be passed to an optimizer such as \R's \textsf{optim} or \textsf{nlminb} with potentially better results than using those functions without bounds.  However, even with known bounds, general purpose optimizers may still miss \mrle\ and regions of high $\logRL$ and they emphasize single points of highest local $\logRL$ while possibly ignoring large adjacent regions where $\logRL$ is nearly as large \citep{hill:1965, hodges:2013}.

We will next find a box $B \supset B_1$ satisfying desideratum \textbf{D1}.  First, though, we pause to note that the region outlined in bold in Figure~(\ref{fig:smallboundingbox}) is a subset of $B_1$ that must also contain \mrle\ by the same reasoning used to prove the previous claim.  But it is not rectangular, hence less convenient than $B_1$, so we don't pursue it further.\\[5pt]

@@

Let $q_2 = (\sigesq{}_2, \sigssq{}_2)$ satisfy $\sigesq{}_2 \ge \sigma_e^{2*}$ and $\sigssq{}_2 \ge \widetilde{\sigma_s^2}$ --- above $B$ and to the right of the $s_z+1$ line.  (Points $q_2$ and $q_3$ in Figure~(\ref{fig:boundingbox2}) both satisfy these criteria.)  The partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) evaluated at $q_2$ are nonpositive for $j \in \{1, \dots, s_z+2\}$; therefore, $\logRL(\sshat) - \logRL(q_2) \ge L - \logRL(\widetilde{q_2})$.  The latter expression is greater than $M$ iff $\logRL(\widetilde{q_2}) < L-M$.  But examination of (\ref{eq:sum}) shows that for any fixed $\sigesq$, $\lim_{\sigssq \rightarrow \infty} \logRLss = -\infty$ and therefore $\widetilde{\sigma_s^2}$ can be chosen large enough so that \textbf{D1} is satisfied.

@@

\subsection{Lines Determine Bounds within Boxes}
\label{sec:boundsinboxes}
\begin{figure}
	\centering
	\includegraphics[width=.5\linewidth]{figs/boxb.pdf}
	\caption{For a box $b$,  the minimum and
	maximum within $b$ of $\logRL_j$
	occur at either the lower left corner, the upper right  corner, or
	on the line.
}	\label{fig:boxb}
\end{figure}
Next we consider the relationship between the $\logRL_j$'s and boxes as depicted in Figure~\ref{fig:boxb}, which shows a small box $b$ and three lines.  For lines such as $j=2$ that lie below or to the left of $b$, the partial derivatives (\ref{eq:drlds}, \ref{eq:drlde}) are nonpositive, so the maximum and minimum of $\logRL_j$ within $b$ are attained at the lower left and upper right corners respectively.  The situation is reversed for lines like $j=3$ that lie above or to the right of $b$: the partial derivatives are nonnegative so the maximum and minimum are attained at the upper right and lower left corners respectively.  For lines like $j=1$ that pass through $b$, the minimum is attained at either the upper right or lower left corner while the maximum is attained on the line.

@@

\section{Examples}
\label{sec:examp}
Throughout Section~\ref{sec:examp} we use $\epsilon = 1$, $M = 7$, $\delta_e = \delta_s = 0$ and $\maxit = 25$.  We used a finite value of $\maxit$ for practicality but, because each example took 25 or fewer iterations to adequately resolve $\log f$, $\maxit = 25$ yields the same results as $\maxit = \infty$.  Computations were performed on an iMac that was new in 2011, having a 2.7 GHz Intel Core i5 processor and 4GB of memory.
\subsection{HMO premiums}
\subsubsection{Introduction to the Data}
Our first example is a traditional linear mixed model previously analyzed in \cite{hodges:98}, \cite{hodges:2013}, and \cite{henn&hodges:2014} who reported a bimodal log posterior density for $(\sigesq,\sigssq)$.  Quoting from \cite{henn&hodges:2014},
\begin{quote}
\itshape
\dots the HMO data set describes 341 HMOs [Health Maintenance Organizations] located in 45 states or similar political jurisdictions.  Each jurisdiction had between 1 and 31 plans with a median of 5 plans.  The data set originally was analysed to assess the cost of moving military retirees and dependents from a Department of Defense health plan to plans serving the US civil service.  \dots.  Specifically, the model is
\begin{equation*}
\begin{split}
	y_{ij} &= \alpha_i + \epsilon_{ij}\\
	\alpha_i &= \varrho_0 + \varrho_1x_{1i} + \varrho_2x_{2i} + \zeta_i,
\end{split}
\end{equation*}
where the fixed effects in $\alpha_i$ include an intercept, jurisdiction-average hospital expenses per admission ($x_{1i}$) and an indicator for plans in New England states ($x_{2i}$).

@@

Figure~\ref{fig:hmoboxes} shows the outlines of the 9490 boxes.  The algorithm did not need to divide the boxes with large $\sigesq$ or $\sigssq$ as finely as those with small $\sigesq$ and $\sigssq$ because they more readily satisfy either $U^b < L-M$ or $U^b - L^b < \epsilon$, so become inactive.  Figure~\ref{fig:hmorll} shows the same boxes on a logarithmic scale, shaded by $L^b$ in each box.  The red dot is the lower left corner of the box that maximizes $L^b$.  Boxes with $U^b \ge L \equiv \max L^b$ are outlined in blue; $(\sshat)$ must lie within the blue region.  
%Figure~\ref{fig:hmorll}  suggests that we can restrict attention to a box determined by $\sigesq\in(300,1000)$ and $\sigssq\in(0,1000)$.  So we rerun the algorithm on that box and with more stringent control parameters:
%\begin{equation*}
%	\maxit=15; \hspace{1cm} \epsilon=1; \hspace{1cm}
%	\delta_e=0; \hspace{1cm} \delta_s=0; \hspace{1cm} M=10
%\end{equation*}
%Table~\ref{table:hmo_HH11_run2} shows the output.  The algorithm needed 12 iterations to move all boxes to the inactive list.  The resulting 37,516 boxes are shown in Figure~\ref{fig:hmo2}.
For comparison, the standard REML analysis using \R's \textcompute{lme} function yields $\sshat \approx (495, 99)$ with 95\% confidence intervals of $(421, 582)$ and $(39, 248)$.
